Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-10 paper/sections/conclusions.tex
added text.

Nevertheless, the proof reconstruction is certainly possible, and
the verification of solutions of \CPL problems can be carried out
successful in \Agda. Our tool is able to translate \Metis
derivations in \TSTP format to deliver \Agda code of the proof,
\change{Better uses \emph{proof-terms}?}
and this proof can be type-checked by \Agda.

>>>>

Nevertheless, the proof-reconstruction is certainly possible, and
the verification of solutions of \CPL problems can be carried out
successful in \Agda. Our tool is able to translate \Metis
derivations in \TSTP format to deliver \Agda code of the proof,
\change{Better uses \emph{proof-terms}?}
and this proof can be type-checked by \Agda.

### 2017-09-10 paper/sections/introduction.tex
[ paper ] section 3 introduction added.
added text.

Proof reconstruction is a hard labor since it depends on the
integration of two complex systems. On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories. A \SAT solver (e.g., \name{zChaff}
\cite{Moskewicz2001} and \name{MiniSat}~\cite{Een2004}) to prove
unsatisfiability of \CNF formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and \name{DepQBF}
\cite{Lonsing2017}) to prove satisfiability and invalidity of
quantified Boolean formulas, a \SMT solver (e.g \name{CVC4}
\cite{Barrett2011}, \name{veriT}~\cite{bouton2009}, and \name{Z3}
\cite{DeMoura2008}) to prove unsatisfiability of formulas from
first-order logic with theories, and a prover for validity of
formulas from first-order logic with equality (e.g., \name{E}
\cite{Schulz:AICOM-2002}, \name{leanCoP}~\cite{Otten2008}, \Metis
\cite{hurd2003first}, \name{SPASS} ~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II} \cite{Benzmuller2008} and \name{Satallax}
\cite{Brown2012}) or intuitionistic logic (e.g., \name{ileanCoP}
\cite{Otten2008}, \name{JProver} \cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam}, \name{Coq}
\cite{coqteam}, \name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems like
operators, logic variables, axioms, and theorems.

A proof reconstruction tool provides such an integration translating
the reply delivered by the prover into the formalism of the proof
assistant. Because the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the
target, the reconstruction turns out in a ``reverse engineering''
task. Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the specification
logic in the \ITP. To begin with the proof reconstruction, it is
necessary to have a proof-object from the \ATP in a consistent
format to work with, that is, a full script describing step-by-step
with exhaustive details and without ambiguities, the derivation to
get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g., \name{CVC4}
\cite{Barrett2011}, \name{SPASS}, and \name{Waldmeister}
\cite{hillenbrand1997}) and a little bit less replied with a proof
in a file format like \TSTP~\cite{sutcliffe2004tstp} (e.g.,
\name{E}, \Metis, \name{Vampire}, and \name{Z3}), \len{LFSC}
\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been
implemented for proof reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013}, \cite{Bohme2011}, \cite{Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related work,
section \ref{sec:related-work}.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows. In section
\ref{sec:metis-language-and-proofs}, we briefly introduce the \Metis
prover. In section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
In section~\ref{sec:conclusions}, we discuss
some limitations, advantages, and disadvantages with other similar
proof reconstruction tools for ending up with some conclusions and the
future work. Finally, we list some related works in proof reconstruction
in section \ref{sec:related-work}.

>>>>

proof-reconstruction is a hard labor since it depends on the
integration of two complex systems. On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories. A \SAT solver (e.g., \name{zChaff}
\cite{Moskewicz2001} and \name{MiniSat}~\cite{Een2004}) to prove
unsatisfiability of \CNF formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and \name{DepQBF}
\cite{Lonsing2017}) to prove satisfiability and invalidity of
quantified Boolean formulas, a \SMT solver (e.g \name{CVC4}
\cite{Barrett2011}, \name{veriT}~\cite{bouton2009}, and \name{Z3}
\cite{DeMoura2008}) to prove unsatisfiability of formulas from
first-order logic with theories, and a prover for validity of
formulas from first-order logic with equality (e.g., \name{E}
\cite{Schulz:AICOM-2002}, \name{leanCoP}~\cite{Otten2008}, \Metis
\cite{hurd2003first}, \name{SPASS} ~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II} \cite{Benzmuller2008} and \name{Satallax}
\cite{Brown2012}) or intuitionistic logic (e.g., \name{ileanCoP}
\cite{Otten2008}, \name{JProver} \cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam}, \name{Coq}
\cite{coqteam}, \name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems like
operators, logic variables, axioms, and theorems.

A proof-reconstruction tool provides such an integration translating
the reply delivered by the prover into the formalism of the proof
assistant. Because the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the
target, the reconstruction turns out in a ``reverse engineering''
task. Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the specification
logic in the \ITP. To begin with the proof-reconstruction, it is
necessary to have a proof-object from the \ATP in a consistent
format to work with, that is, a full script describing step-by-step
with exhaustive details and without ambiguities, the derivation to
get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g., \name{CVC4}
\cite{Barrett2011}, \name{SPASS}, and \name{Waldmeister}
\cite{hillenbrand1997}) and a little bit less replied with a proof
in a file format like \TSTP~\cite{sutcliffe2004tstp} (e.g.,
\name{E}, \Metis, \name{Vampire}, and \name{Z3}), \len{LFSC}
\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been
implemented for proof-reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013}, \cite{Bohme2011}, \cite{Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related work,
section \ref{sec:related-work}.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows. In section
\ref{sec:metis-language-and-proofs}, we briefly introduce the \Metis
prover. In section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
In section~\ref{sec:conclusions}, we discuss
some limitations, advantages, and disadvantages with other similar
proof-reconstruction tools for ending up with some conclusions and
the future work. Finally, we list some related works in proof
reconstruction in section \ref{sec:related-work}.

### 2017-09-10 paper/sections/metis-language-and-proofs.tex
added text.

\Metis is an automatic theorem prover written in \len{Standard ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof reconstruction
with \Metis, some of these systems are the following.

@@

This lack of the \Metis' documentation is an important matter for
us in the proof reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify.
For example, we believe by applying these last
inferences, \Metis summarized many essential steps to deduce a
formula from a list a list of formulas as we see in the following
example.
\improvement{Add an example here to show a deduction where \Metis do
a lot between two steps.}
That inference' behavior has become a drawback since, our approach
to reconstruct proofs needs take into account any assumption or
consideration as ordering for literals, application of
simplification theorems, reductions, among other valid
transformations.

>>>>

\Metis is an automatic theorem prover written in \len{Standard ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof-reconstruction
with \Metis, some of these systems are the following.

@@

This lack of the \Metis' documentation is an important matter for
us in the proof-reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify.
For example, we believe by applying these last
inferences, \Metis summarized many essential steps to deduce a
formula from a list a list of formulas as we see in the following
example.
\improvement{Add an example here to show a deduction where \Metis do
a lot between two steps.}
That inference' behavior has become a drawback since, our approach
to reconstruct proofs needs take into account any assumption or
consideration as ordering for literals, application of
simplification theorems, reductions, among other valid
transformations.

### 2017-09-10 paper/sections/related-work.tex
[ paper ] appendix added.
added text.

\name{Sledgehammer} is a tool for \name{Isabelle} proof assistant
\cite{paulson1994isabelle} that provides a full integration of
automatic theorem provers including \ATPs (see, for example,
\cite{meng2006automation}, \cite{blanchette2013extending} and
\cite{Fleury2014}) and \SMT solvers (see, for instance,
\cite{hurlin07practical}, \cite{bohme2010}, \cite{blanchette2013extending}, and \cite{Fleury2014}) with
\name{Isabelle/HOL} \cite{nipkow2002isabelle}, the specialization of
\name{Isabelle} for higher-order logic. Sultana, Benzm{\"{u}}ller,
and Paulson~\cite{Een2004} integrates \name{Leo-II} and
\name{Satallax}, two theorem provers for high-order logic with
\name{Isabelle/HOL} proposing a modular proof reconstruction work
flow.

@@

\name{Waldmeister} is an automatic theorem prover for unit
equational logic \cite{hillenbrand1997}. Foster and Struth
\cite{foster2011integrating} integrate \name{Waldmeister} into \Agda
\cite{agdateam}. This integration requires a proof reconstruction
step, but authors' approach is restricted to pure equational logic
--also called identity theory~\cite{humberstone2011}-- that is,
first-order logic with equality but no other predicate symbols and
no functions symbols~\cite{appel1959}.

>>>>

\name{Sledgehammer} is a tool for \name{Isabelle} proof assistant
\cite{paulson1994isabelle} that provides a full integration of
automatic theorem provers including \ATPs (see, for example,
\cite{meng2006automation}, \cite{blanchette2013extending} and
\cite{Fleury2014}) and \SMT solvers (see, for instance,
\cite{hurlin07practical}, \cite{bohme2010},
\cite{blanchette2013extending}, and \cite{Fleury2014}) with
\name{Isabelle/HOL} \cite{nipkow2002isabelle}, the specialization of
\name{Isabelle} for higher-order logic. Sultana, Benzm{\"{u}}ller,
and Paulson~\cite{Een2004} integrates \name{Leo-II} and
\name{Satallax}, two theorem provers for high-order logic with
\name{Isabelle/HOL} proposing a modular proof-reconstruction work
flow.

@@

\name{Waldmeister} is an automatic theorem prover for unit
equational logic \cite{hillenbrand1997}. Foster and Struth
\cite{foster2011integrating} integrate \name{Waldmeister} into \Agda
\cite{agdateam}. This integration requires a proof-reconstruction
step, but authors' approach is restricted to pure equational logic
--also called identity theory~\cite{humberstone2011}-- that is,
first-order logic with equality but no other predicate symbols and
no functions symbols~\cite{appel1959}.
